---
title: Bibliometrics with OpenAlex
teaching: 45
exercises: 15
source: Rmd
---

::: objectives
- Decide whether OpenAlex represents the thing a patron is asking about: an author, a
  research center with its own entry, or a department with no entry of its own
- Query the OpenAlex API from R using `openalexR`
- Apply prior data-cleaning skills to nested/list-column API data
- Reuse `ggplot2` patterns from Data Visualization with ggplot2 on a new dataset
- Knit a short, honest bibliometric snapshot from a starter report template
:::

::: questions
- A patron asks a bibliometrics question. What do I need to figure out before I can even
  query OpenAlex?
- How do I pull publication data for an author or institution into R, without waiting forever
  or hitting a wall?
- How is API data different from a clean CSV, and how do I tidy it?
- How do the visualization and reporting skills from earlier episodes transfer to new data?
- What can this kind of snapshot not tell a patron?
:::

```{r, include=FALSE}
source("files/get_openalex_data.R")
library(openalexR)
library(dplyr)
library(tidyr)
library(ggplot2)
pkp_works <- readRDS("data/pkp-works-2026-09-20.rds")
```

## Setup

This episode needs one new package. Install it the same way you installed `tidyverse` back
in Before We Start:

```{r, eval=FALSE}
install.packages("openalexR")
```

`dplyr`, `tidyr`, and `ggplot2` are packages you already installed and used in earlier
episodes, nothing new there:

```{r, eval=FALSE}
# Already installed and loaded from earlier episodes:
# library(dplyr)
# library(tidyr)
# library(ggplot2)

# New for this episode:
library(openalexR)
```

## A patron comes to the desk

You're at the reference desk, virtual or in person, and the director of a research center
asks for an exploratory snapshot: how has the center's publication output and research focus
changed over time? This episode teaches you to put together an answer with
[OpenAlex](https://openalex.org/) and R, the same import, clean, visualize, and report
workflow from the last four episodes, applied to scholarly metadata instead of circulation
data.

OpenAlex is a free, open catalog of the world's research (works, authors, institutions), with
a CC0 license and no account required for casual use. Before you write any code, the first
real question is a reference-interview question, not a technical one: does OpenAlex represent
what this patron is asking about as its own distinct entity?

::: callout

## Accessing OpenAlex

OpenAlex works without an account for the kind of casual, occasional use this workshop needs.
A free OpenAlex API key raises your daily request budget if you ever need more than that, but
you won't need one for anything here. Don't put a real API key, or anyone's personal email,
into lesson code, a saved data file, or anything you commit to a repository.

:::

## Is this a person, a center, or a department?

A bibliometrics request usually falls into one of three cases, and each one changes how much
work you're signing up for.

1. **An author.** One person's own publication record. Usually queryable once you've confirmed
   you have the right person. Common names collide, so check an ORCID if the patron has one.
2. **An organization with its own OpenAlex entry.** Could be a university, a research center,
   a museum, a nonprofit, anything with a distinct identity in OpenAlex. Small, checkable, and
   often fetchable live during a workshop. The Public Knowledge Project, used below, is this
   case.
3. **A department with no entry of its own.** Very common ask ("how has my department's output
   changed?"), and OpenAlex usually has no clean answer, because most departments aren't a
   distinct entity in OpenAlex the way a whole university or research center can be. Answering
   it for real would need a verified roster of the department's authors, their membership
   dates, and a rule for handling joint appointments and cross-affiliations. That's a real
   project on its own, not something to solve in one workshop exercise.

A whole university is usually its own entity too, but that doesn't make it a good fit for a
quick workshop demo: it's normally far too large to fetch or inspect live.

```{r, eval=FALSE}
# Not evaluated. Illustrates scale only.
oa_fetch(
  entity = "works",
  authorships.institutions.id = "SOME_LARGE_UNIVERSITY_ID",
  count_only = TRUE
)
# A large research university can have tens of thousands of works indexed.
# That's too many to fetch or inspect live in a workshop, and it's exactly
# why the worked example below uses something smaller.
```

## Resolving the entity: two worked cases

Two short worked examples below show what searching, inspecting, and confirming an entity
involves: once for an organization, once for a person.

### Case 1: an organization

The instructor demo below profiles the Public Knowledge Project (PKP), a nonprofit that builds
free, open-source scholarly-publishing software, including Open Journal Systems, which many
libraries run to publish their own journals. It's a deliberately manageable example: a
distinct OpenAlex organization, a small and workshop-sized set of works, and direct relevance
to libraries, publishing, and scholarly communication. Most university departments won't be
this straightforward either way, which is the whole point of the discussion above.

```{r, eval=FALSE}
# Search first to confirm you have the right organization. Search terms can
# be ambiguous (similarly named organizations, regional branches).
institution_matches <- oa_fetch(
  entity = "institutions",
  search = "Public Knowledge Project"
)

institution_matches |>
  select(display_name, id, ror, works_count)
```

A single match: Public Knowledge Project, ROR `ror.org/05ek4tb53`. Its own `works_count` field
reports 148, a running total OpenAlex keeps for the organization overall. That's not the same
number as what a specific, filtered query returns, which is why this episode always trusts a
real query result over a summary field like this one.

```{r}
institution_id <- "I4387153203"
```

The line below is what produced this episode's data (details in
`data/pkp-works-2026-09-20-PROVENANCE.md`, including the retrieval date and package version),
but it's shown here as a comment, not run live during the workshop:

```{r, eval=FALSE}
# Shown for provenance, not executed during the workshop.
# pkp_works <- oa_fetch(
#   entity = "works",
#   authorships.institutions.id = institution_id,
#   from_publication_date = "1998-01-01"
# )
```

The date filter above is doing real work, not only narrowing scope: an unfiltered fetch
includes two records dated 1969, decades before PKP was founded in 1998. That kind of outlier
is common enough in API data that it's worth checking for before trusting a count.

`readRDS()` loads R's own native saved-object format. Unlike a CSV, an `.rds` file preserves
list-columns exactly as they were, which matters for the nested data below.

```{r}
pkp_works <- readRDS("data/pkp-works-2026-09-20.rds")
nrow(pkp_works)
```

`r nrow(pkp_works)` works is small enough to fetch live in a workshop session without
rate-limit or runtime trouble, worth checking before committing to a bigger query. This is an
exploratory snapshot of what OpenAlex had indexed as of the retrieval date, not a complete or
authoritative evaluation of PKP's output.

### Case 2: a person

What if the patron is asking about a person instead? This episode's author, Tim Dennis, agreed
to let his own public OpenAlex record stand in as the example, so this case is a live worked
example rather than a hypothetical.

```{r, eval=FALSE}
author_matches <- oa_fetch(
  entity = "authors",
  search = "Tim Dennis"
)

author_matches |>
  select(display_name, id, orcid, works_count, last_known_institutions)
```

That search returns more than a dozen different people who share this name. A name alone is
never a unique identifier: the next step is checking each candidate's ORCID, affiliation, or a
known work against what you already know about the person you mean.

One candidate has ORCID `0000-0001-6632-3812`. ORCID's own public record for that ID lists
"Director, Data Science Center, UCLA Library," starting in 2017, matching Tim Dennis's current
role, along with earlier positions at UC San Diego and UC Berkeley Libraries. That's a strong
match on affiliation and identity.

```{r}
author_id <- "A5101548578"
```

Here's the twist worth sitting with: a correct ORCID match doesn't guarantee a clean OpenAlex
author record. Pull up the works attached to this ID and you'll find titles like "Epipolar line
estimation and rectification" and "Image sequence stabilisation based on DFT filtering,"
image-processing research with no connection to library data science. This is where the third
disambiguation signal earns its keep: a known work. "Top 10 FAIR Data & Software Things" (2019)
is recognizably his, a short, citable piece from the open-science community he works in.
Spotting one title you recognize among a run of unrelated ones is exactly how you'd notice, in
practice, that a record has been merged rather than trusting the ORCID alone and moving on.
OpenAlex has combined this ORCID with a different researcher who happens to share the same
name. The ID above is still the right one to have selected, since it's the record tied to the
verified ORCID, but that doesn't make a raw works count from it trustworthy, so this episode
stops here rather than running one. Sometimes the answer to "how many works does this person
have" is "not a number this particular record can give you cleanly."

::: callout

Tim Dennis, this lesson's author, has agreed to the use of his real, public OpenAlex and ORCID
records here. OpenAlex data can contain omissions, duplicates, incorrect affiliations, or
attribution errors like the one above. This example is about disambiguation, not an evaluation
of anyone's research.

:::

::: instructor

## Substituting your own record

If you're comfortable using your own OpenAlex author record instead of Tim Dennis's for this
demo, you can, but test it before the workshop: search for yourself, and confirm the match
using an ORCID, affiliation, or a known publication, the same as above. If you don't have a
usable record, or would rather not use your own, the Tim Dennis example works fine as-is.

Don't display or interpret your own citation counts in front of the room, for the same reason
this episode doesn't for Tim Dennis. And don't ask learners to search for or disclose their own
publication or citation record: having no OpenAlex record at all is normal and says nothing
about anyone's expertise.

:::

::: challenge

## Challenge: Identify and fetch your own

Imagine a patron has asked you their bibliometrics question. First, which of the three cases
above is it: an author, an organization with its own entry, or a department without one? Then
work through the same steps as the worked cases above for your own choice: search, inspect the
candidates, and save a verified ID.

If you'd rather use your own name or [ORCID](https://orcid.org/) than a hypothetical patron,
that's fine. Nobody needs to search for or share their own publication record if they'd rather
not, and a thin or empty result says something about OpenAlex's coverage, not about anyone's
work.

**Give yourself about five minutes.** If your search is ambiguous, returns nothing usable, or
you're still not confident in a match after that, stop and use the fallback below instead. None
of the remaining exercises depend on your own fetch having worked.

```{r, eval=FALSE}
my_works <- pkp_works
```

::: solution

Institution path:

```{r, eval=FALSE}
institution_matches <- oa_fetch(
  entity = "institutions",
  search = "your institution's name"
)

institution_matches |>
  select(display_name, id, ror, works_count)

institution_id <- "PASTE_THE_ID_YOU_CONFIRMED"

oa_fetch(
  entity = "works",
  authorships.institutions.id = institution_id,
  count_only = TRUE
)

# Only fetch works if that count feels workshop-sized:
my_works <- oa_fetch(
  entity = "works",
  authorships.institutions.id = institution_id,
  from_publication_date = "2020-01-01"
)
```

Author path:

```{r, eval=FALSE}
author_matches <- oa_fetch(
  entity = "authors",
  search = "the name you're checking"
)

author_matches |>
  select(display_name, id, orcid, works_count, last_known_institutions)

author_id <- "PASTE_THE_ID_YOU_CONFIRMED"

oa_fetch(
  entity = "works",
  authorships.author.id = author_id,
  count_only = TRUE
)

my_works <- oa_fetch(
  entity = "works",
  authorships.author.id = author_id,
  from_publication_date = "2020-01-01"
)
```

Both return a data frame, one row per work, with the same columns shown below on the PKP data. For institutions, double-check the name and, if available, its ROR ID against your search
result. For authors, an ORCID is the most reliable disambiguator, though as the worked example
above shows, even that isn't a complete guarantee. If you didn't manage a confident match,
`my_works <- pkp_works` is a completely fine place to be for the rest of this episode.

:::
:::

::: discussion

## Optional: why query this directly?

*Skip this first if you're running short on time.*

A patron at the desk needs something by end of day. OpenAlex is free, no account required, CC0
licensed. What's different about reaching for this versus a vendor tool's export, like Scopus
or Web of Science? Think about access, transparency, coverage, name disambiguation, and whether
someone else could reproduce your result.

:::

## Clean and reshape the data

`pkp_works` is a data frame, one row per work, the same shape you already know. What's
different from `books.csv` is that some of its columns aren't simple values: they're
**list-columns**, each cell holding a small data frame of its own.

```{r}
glimpse(pkp_works)
```

That's a lot of columns at once. Don't try to take in all 44: focus specifically on `id`,
`display_name`, and `publication_year` (flat, ready to use as-is), and at `authorships` and
`topics` (list-columns, need more work first, covered next).

To work with a list-column, unnest it into its own tidy table. A first attempt at unnesting
`topics` hits a real, common snag:

```{r, eval=FALSE}
pkp_works |>
  select(id, display_name, topics) |>
  unnest(topics)
```

```
Error in `unnest()`:
! Can't duplicate names between the affected columns and the original data.
✖ These names are duplicated:
  ℹ `id` and `display_name`, from `topics`.
ℹ Use `names_sep` to disambiguate using the column name.
```

The work itself has an `id` and `display_name`, and so does every topic attached to it,
so `unnest()` refuses to guess which is which. Fix it with `names_sep`, which prefixes the
unnested columns:

```{r}
pkp_topics <- pkp_works |>
  select(id, display_name, publication_year, topics) |>
  unnest(topics, names_sep = "_")

nrow(pkp_topics)
```

`r nrow(pkp_works)` works became `r nrow(pkp_topics)` rows: OpenAlex tags each work with
several topics, at several levels of a hierarchy (`topic`, `subfield`, `field`, `domain`), so
one work now appears once per topic-level combination. That row count jump is expected, not a
bug, but it's worth noticing before it surprises you downstream, including a few sections from
now, when counting "works per field" needs one more step because of it. It's also why the
report step later calculates totals from `pkp_works`, one row per work, and not from this
unnested table, which would badly overcount.

::: challenge

## Optional stretch: unnest a second field

*If you have time left, try the same pattern on `authorships` instead of `topics`. Expect a
bigger row jump: a work with several authors becomes several rows, one per author.*

::: solution

```{r, eval=FALSE}
pkp_authors <- pkp_works |>
  select(id, display_name, publication_year, authorships) |>
  unnest(authorships, names_sep = "_")
```

:::
:::

## Visualize

The same `ggplot2` patterns from Data Visualization with ggplot2 transfer directly.

```{r, fig.alt="Bar chart of Public Knowledge Project publications by year, showing a handful of works per year from 2005 through 2024, then a sharp jump to 20 works in 2025 and 28 in 2026"}
pkp_works |>
  count(publication_year) |>
  ggplot(aes(x = publication_year, y = n)) +
  geom_col() +
  labs(x = "Publication year", y = "Works",
       title = "Public Knowledge Project publications per year")
```

Output stays low and uneven for most of the range, then jumps sharply in the last two years,
which is worth naming out loud as a real pattern rather than glossing over: it could reflect a
genuine increase in output, a change in what OpenAlex indexes, or both, and a patron reading
this chart would reasonably ask which one it is.

::: challenge

## Challenge: Transfer the pattern

Reproduce this chart for your own institution, author, or the cached data, starting from
your notes from Data Visualization with ggplot2 rather than being handed new code.

::: solution

```{r, eval=FALSE}
my_works |>
  count(publication_year) |>
  ggplot(aes(x = publication_year, y = n)) +
  geom_col() +
  labs(x = "Publication year", y = "Works")
```

Adaptation usually needed: your date range and count scale will differ, and a small dataset may
have fewer bars than PKP's, which is real signal, not an error.

:::
:::

The instructor can also demonstrate a second chart type, filtering to the top level of the
topic hierarchy. One real wrinkle first: a single work can carry more than one topic mapped to
the same field, so counting rows in `pkp_topics` would count some works more than once. Count
distinct works per field instead:

```{r, fig.alt="Horizontal bar chart of Public Knowledge Project's top research fields by number of distinct works, led by Computer Science and Decision Sciences"}
pkp_topics |>
  filter(topics_type == "field") |>
  distinct(id, topics_display_name) |>
  count(topics_display_name, sort = TRUE) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = n, y = reorder(topics_display_name, n))) +
  geom_col() +
  labs(x = "Distinct works tagged with this field", y = NULL,
       title = "PKP's top research fields")
```

Computer Science and Decision Sciences lead, well ahead of Social Sciences and Arts and
Humanities.

::: challenge

## Optional stretch: pick your own question

*Skip this if you're running short on time. It isn't required for the report below.*

Try the top-fields chart above on your own data, or try a citation-count distribution
(`geom_histogram()` on `cited_by_count`) instead. There's no single right answer: the point is
picking whichever chart answers the patron's question.

:::

## Report back to the patron

The point of all this is a report someone else can understand, reinforcing the reproducibility
framing from Reproducible Reports with Quarto, not hand-copied numbers.

```{r}
n_works <- nrow(pkp_works)
year_range <- range(pkp_works$publication_year, na.rm = TRUE)
```

A short, honest snapshot might say:

> As of the retrieval date recorded in this project, the Public Knowledge Project has
> `r n_works` works indexed by OpenAlex, spanning `r year_range[1]` to `r year_range[2]`, most
> often tagged with Computer Science and Decision Sciences fields.

That's deliberately modest language: an indexed count from one open database, not a claim
about everything PKP has ever published or built. Publication and citation
counts are a partial
picture, not a complete measure of research quality, and a current year's total is usually an
undercount because indexing takes time to catch up.

::: challenge

## Challenge: Capstone report

Use the [starter template](files/patron-report-template.qmd) to knit a short report as if you
were about to hand it to the patron from the frame story: written for them, not for yourself.
If you used your own author record earlier, you can report on the cached data instead of
sharing anything personal.

A finished report should:

- name the author or institution it's about
- give the retrieval or fixture date
- say the works are "indexed by OpenAlex," not a complete count
- give a date range and total work count, calculated from one row per work
- include one clearly labeled plot
- state one finding in plain language
- name at least one real limitation
- render without errors

Swap with a neighbor playing the patron. Can they answer their own question from your report
alone, without you explaining it?

::: solution

There's no fixed answer key. The real check is whether a "patron" reading it cold gets their
answer. If they can't, the summary sentence probably needs to say more, or the plot needs a
clearer title or label.

:::
:::

## Wrap-up discussion

What would you still want to know about this institution or author that a snapshot like this
alone can't tell you?

::: keypoints
- OpenAlex works without an account for casual use like this, and doesn't need a personal
  email or API key for a workshop-scale query.
- Deciding whether a patron's request is an author, an organization with its own entry, or a
  department with no entry of its own is a reference-interview skill, not only an API choice.
- API responses often include list-columns. `tidyr::unnest(..., names_sep = "_")` flattens them
  while avoiding name collisions with the outer table.
- Unnesting a list-column multiplies rows, one per nested value, so totals should come from the
  original one-row-per-work table, not the unnested one.
- OpenAlex's current subject field is `topics`, not the older, unmaintained `concepts` field.
- A verified ORCID confirms you have the right person, but not that OpenAlex's own record for
  them is clean. Check actual works before trusting a count.
- The same `ggplot2` and reporting skills from earlier episodes transfer directly to new,
  API-sourced data, and any report should say plainly what it can't tell you.
:::
