---
title: "Bibliometric snapshot"
format: html
---

<!-- Fill in each TODO. Aim for 10-15 minutes. -->

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

# TODO: point this at your own fetched data, or the cached PKP fixture
# my_works <- readRDS("data/pkp-works-2026-09-20.rds")
```

## Who and when

<!-- TODO: name the institution or author, and the retrieval or fixture date -->

This snapshot covers works indexed by OpenAlex as of TODO-DATE.

```{r}
# TODO: total works and date range, calculated from one row per work
n_works <- nrow(my_works)
year_range <- range(my_works$publication_year, na.rm = TRUE)
```

## What the data shows

```{r}
# TODO: one clearly labeled plot (reuse your publications-per-year code)
```

TODO: one or two sentences stating the main finding, written for the patron,
not for yourself.

## Limitations

TODO: name at least one coverage or interpretation limitation (for example:
current-year totals may be incomplete, or author/affiliation disambiguation
can be wrong).
